\chapter*{Lời Kết}
\addcontentsline{toc}{chapter}{Lời Kết}
\markboth{Lời Kết}{Lời Kết}

\begin{truyen}{Lời Kết}{Closing Words}
\chuhoa{S}{ống chắt chiu chưa bao giờ là chuyện đẹp trên lời nói mà xấu trong thực tế. Trái lại, đó thường là dáng đứng thầm lặng của những người hiểu đời không dễ.}
\textit{(Living frugally has never been something beautiful only in words and ugly in practice. On the contrary, it is often the quiet posture of those who understand that life is not easy.)}

Người biết để dành không phải vì họ không dám sống. Họ chỉ không muốn ngày mai của con cái, cha mẹ hay chính mình bị đẩy vào khốn đốn vì một hôm tiêu tiền theo cảm xúc. Họ hiểu rằng một mái nhà bình yên không được dựng bằng vài quyết định hoành tráng, mà bằng rất nhiều lần biết dừng lại trước một món không cần.
\textit{(People who know how to save do not live cautiously because they are afraid of life. They simply do not want tomorrow for their children, parents, or themselves to be thrown into distress because of one emotional day of spending. They understand that a peaceful home is not built by a few grand decisions, but by many moments of stopping before unnecessary purchases.)}

Nếu cuốn sách này có giá trị gì, thì đó là nhắc người đọc rằng giữ tiền cho đúng nhiều khi cũng là giữ phẩm giá, giữ bình an, giữ quyền được thở chậm hơn khi biến cố đến. Mong rằng sau trang cuối, mỗi người sẽ tự viết thêm cho mình những chương mới bằng chính nếp sống hằng ngày.
\textit{(If this book has any value, it is in reminding the reader that keeping money wisely is often also a way of keeping dignity, peace, and the right to breathe more calmly when hardship arrives. May each reader continue writing new chapters through daily habits after the final page.)}
\end{truyen}

\begin{baihoc}
Không ai đổi đời chỉ vì đọc xong một cuốn sách. Nhưng người biết lặp lại một điều đúng mỗi ngày có thể cứu cả gia đình mình theo thời gian.
\textit{(No one changes life merely by finishing a book. But someone who repeats one right thing every day may save an entire family over time.)}
\end{baihoc}

\begin{ghinhoanh}
\textbf{Short English to remember:}

Small discipline, repeated daily,

can become a shelter for a whole family.
\end{ghinhoanh}

\ngancach